### 2025-07-10 paper/quarto_aaai/paper.tex
minor adjustments to 3.1

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
generated through gradient descent over \(T\) iterations as originally
proposed by \citet{wachter2017counterfactual}. In broad terms, searching
for CEs using gradient descent entails optimizing some form of an
objective that balances (1) the classification loss
\(\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y})\), and (2) one
or more penalty terms \(\lambda_{i}\text{cost}_{i}(\cdot)\). The exact
specification of these penalties induces various properties in the
counterfactual outcomes, and tends to be the key feature that
distinguishes various gradient-based ``generators'' or ``explainers''
(CE algorithms) from the literature \citep{altmeyer2023explaining},
including all generators used in our experiments. We refer the reader to
the supplementary appendix for details.

For our purposes, we let \(T\) vary and consider the search to be
converged as soon as the predicted probability for the target class has
reached a pre-determined threshold, \(\tau\):
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
where \(\mathcal{S}\) is the softmax function.

CT adapts gradient-based CE search already in training to generate
\(\mathbf{x}^\prime\), the perturbed versions of the factual samples
\(\mathbf{x}\). We use the term \emph{nascent} to denote counterfactuals
for which the search has not converged, \(\mathbf{x}_{t\leq T}^\prime\).
Initially, when the perturbations are small, nascent counterfactuals
serve as AEs. Over multiple iterations, they move towards the decision
boundary and turn into \emph{mature} counterfactuals
\(\mathbf{x}_T^\prime\) when they cross into the target class (becoming
CEs) or when \(t > T\).

>>>>

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
generated through gradient descent over \(T\) iterations as originally
proposed by \citet{wachter2017counterfactual}. In broad terms, searching
for CEs using gradient descent entails optimizing some form of an
objective that balances (1) the classification loss
\(\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y})\), and (2) one
or more penalty terms \(\lambda_{i}\text{cost}_{i}(\cdot)\). The exact
specification of these penalties induces various properties in the
counterfactual outcomes, and tends to be the key feature that
distinguishes various gradient-based ``generators'' or ``explainers'' in
the literature \citep{altmeyer2023explaining}, including all generators
used in our experiments. We refer the reader to the supplementary
appendix for details.

CT adopts gradient-based CE search during training to generate
on-the-fly model explanations \(\mathbf{x}^\prime\) for training
samples. We use the term \emph{nascent} to denote counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) that are not yet valid where \(t\)
indicates the last iteration before the label is flipped. We store and
use these interim counterfactuals as adversarial examples. Conversely,
we consider counterfactuals \(\mathbf{x}_T^\prime\) as \emph{mature}
explanations if they have either exhausted all \(T\) iterations or
converged in terms reaching a pre-specified threshold, \(\tau\), for the
predicted probability of the target class:
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
where \(\mathcal{S}\) is the softmax function.

### 2025-07-10 paper/sections/method.qmd
minor adjustments to 3.1

Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual generated through gradient descent over $T$ iterations as originally proposed by @wachter2017counterfactual. In broad terms, searching for CEs using gradient descent entails optimizing some form of an objective that balances (1) the classification loss $\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y})$, and (2) one or more penalty terms $\lambda_{i}\text{cost}_{i}(\cdot)$. The exact specification of these penalties induces various properties in the counterfactual outcomes, and tends to be the key feature that distinguishes various gradient-based "generators" or "explainers" (CE algorithms) from the literature [@altmeyer2023explaining], including all generators used in our experiments. We refer the reader to the supplementary appendix for details.

For our purposes, we let $T$ vary and consider the search to be converged as soon as the predicted probability for the target class has reached a pre-determined threshold, $\tau$: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.

CT adapts gradient-based CE search already in training to generate $\mathbf{x}^\prime$, the perturbed versions of the factual samples $\mathbf{x}$. We use the term *nascent* to denote counterfactuals for which the search has not converged, $\mathbf{x}_{t\leq T}^\prime$. Initially, when the perturbations are small, nascent counterfactuals serve as AEs. Over multiple iterations, they move towards the decision boundary and turn into *mature* counterfactuals $\mathbf{x}_T^\prime$ when they cross into the target class (becoming CEs) or when $t > T$.

>>>>

Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual generated through gradient descent over $T$ iterations as originally proposed by @wachter2017counterfactual. In broad terms, searching for CEs using gradient descent entails optimizing some form of an objective that balances (1) the classification loss $\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y})$, and (2) one or more penalty terms $\lambda_{i}\text{cost}_{i}(\cdot)$. The exact specification of these penalties induces various properties in the counterfactual outcomes, and tends to be the key feature that distinguishes various gradient-based "generators" or "explainers" in the literature [@altmeyer2023explaining], including all generators used in our experiments. We refer the reader to the supplementary appendix for details.

CT adopts gradient-based CE search during training to generate on-the-fly model explanations $\mathbf{x}^\prime$ for training samples. We use the term *nascent* to denote counterfactuals $\mathbf{x}_{t\leq T}^\prime$ that are not yet valid where $t$ indicates the last iteration before the label is flipped. We store and use these interim counterfactuals as adversarial examples. Conversely, we consider counterfactuals $\mathbf{x}_T^\prime$ as *mature* explanations if they have either exhausted all $T$ iterations or converged in terms reaching a pre-specified threshold, $\tau$, for the predicted probability of the target class: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.
